### 2016-02-29 paper/icfp16-submission/semantics.tex
start adding appendix with proofs
abakst comments

\paragraph{Holes}
\label{sec:holes}
%
Recall that a key challenge in our setting is to find witnesses
that are meaningful and do not arise from choosing values from
irrelevant types.
%
We solve this problem be equipping our term language with a notion
of a \emph{hole}, written \ehole, which represents \emph{unconstrained}
values that may replaced with values from \emph{any} type.
%
Dually, holes may also appear in types, written \thole, where they may
be thought of as type variables that we will not generalize over.
%

\paragraph{Substitutions}
%
Our semantics ensure the generality of witnesses by incrementally
\emph{refining} holes, filling in just as much information as is
needed locally to make progress (inspired by the manner in
which SmallCheck uses lazy evaluation~\cite{runciman_smallcheck_2008}).
%
We track how the holes are incrementally filled in, by using
(value) \emph{substitutions} that map holes (denoted by their index)
to values which may, in turn, contain holes.
%
The substitutions let us ensure that we instantiate each hole
with the same (partially defined) value, regardless of the multiple
contexts in which the hole appears, and hence, ensures we can
report a concrete (and general) witness for any (dynamically)
discovered type errors.

@@

\paragraph{Narrowing Types} The procedure $\force{v}{t}{\vsu}{\tsu}$,
defined in Figure~\ref{fig:narrow}, takes as input a value $v$ and
refines it to have the type $t$ by yielding a pair of either the same
value and value substitution, or yields the stuck term if no such
refinement is possible. In the case where $v$ is a hole, it first checks
in the given $\vsu$ to see if the hole has already been instantiated
and, if so, returns the existing instantiation.
%
\begin{figure*}[t]
$$
\begin{array}{lcl}
%% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
\forcesym                  & ::     & (v, t, \vsu, \tsu) \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu} \\
\force{v}{\thole}{\vsu}{\tsu}  & \defeq & \hspace{-1ex}
\begin{cases}
  \force{v}{\subst{\tsu}{\thole}}{\vsu}{\tsu} & \mbox{if}\ \thole \in dom(\tsu) \\
  \triple{v}{\vsu}{\extendsu{\tsu}{\thole}{\typeof{v}}} & \mbox{otherwise} \\
\end{cases} \\
\force{\vhole{t_1}}{t_2}{\vsu}{\tsu} & \defeq & \hspace{-1ex}
\begin{cases}
  \triple{v}{\vsu}{\tsu''}    & \mbox{if } v = \lookupsu{\vsu}{\ehole},
                                         \tsu' = \unify{t_1}{t_2}{\tsu},
                                         \tsu'' = \unify{\subst{\tsu'}{t_1}}
                                                        {\typeof{v}}{\tsu'} \\
  \triple{\stuck}{\vsu}{\tsu} & \mbox{if } v = \lookupsu{\vsu}{\ehole} \\
  \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu'} & \mbox{if}\ \tsu' = \unify{t_1}{t_2}{\tsu}, v = \gen{\subst{\tsu'}{t_1}}\\
\end{cases} \\
% \begin{cases}
%   \triple{\lookupsu{\vsu}{\ehole}}{\vsu}{\tsu} & \mbox{if}\ \ehole \in dom(\vsu), \hastype{\lookupsu{\vsu}{\ehole}}{t} \\
%   \triple{\stuck}{\vsu}{\tsu}                  & \mbox{if}\ \ehole \in dom(\vsu) \\
%   \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu}   & v = \gen{t}\\
% \end{cases} \\
 % \begin{array}{l}
    % \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \triple{\vsu(i)}{\vsu}\ \mathtt{else} \\
         % \elet{v}{\gen{t}}{\triple{v}{\ehole{i} \mapsto v}}
  % \end{array} \\
\force{n}{\tint}{\vsu}{\tsu}     & \defeq & \triple{n}{\vsu}{\tsu} \\
\force{b}{\tbool}{\vsu}{\tsu}    & \defeq & \triple{b}{\vsu}{\tsu} \\
\force{\efun{x}{e}}{\tfun}{\vsu}{\tsu} & \defeq & \triple{\efun{x}{e}}{\vsu}{\tsu} \\
\force{\vleaf{t_1}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vleaf{t_1}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{t_1}{t_2}{\tsu} \\
\force{\vnode{t_1}{v_1}{v_2}{v_3}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vnode{t_1}{v_1}{v_2}{v_3}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{t_1}{t_2}{\tsu} \\
\force{v}{t}{\vsu}{\tsu} & \defeq & \triple{\stuck}{\vsu}{\tsu}
\end{array}
$$
\caption{Narrowing values}
\label{fig:narrow}
\end{figure*}
%
While a hole may map to a value that \emph{contains} another hole, \eg a
lambda or a tree, it may not map \emph{directly} to another hole, so in
the second case above we do not need to \forcesym\ the result of the
substitution, the sub-hole will be narrowed when the context demands it.

\paragraph{Generating Values} The (non-deterministic) procedure $\gen{t}$
takes as input a type \texttt{t} and returns a value of that type. For base
types, the procedure returns an arbitrary value. For functions, it returns
a lambda with a \emph{new} hole denoting the return value, and unconstrained
types (denoted by $\thole$) yield fresh, unconstrained values
(denoted by $\ehole$).

@@

\paragraph{Steps and Traces}
\input{operational.tex}
%
% WRW notes that Figure 4 does not seem to handle recursion (it's not clear
% how the let rule would work for something "let rec"-y, and there's not
% function call rule). I only mention this because I can imagine a reviewer
% wondering about your ability to generate good witnesses for function
% types. This could likely be addressed in text, by a forward reference to
% Section 3.4 where higher-order functions are handled, without changing
% any of the formalisms at the last minute.
Figure~\ref{fig:operational} describes the small-step contextual
reduction semantics for \lang.
%
A \emph{closed expression} is a triple $\triple{e}{\vsu}{\tsu}$ of an expression $e$,
a value substitution $\vsu$, and a type substitution $\tsu$.
%
We write $\step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$ if the closed
expression $\triple{e}{\vsu}{\tsu}$ transitions in a \emph{small step} to
$\triple{e'}{\vsu'}{\tsu'}$.
%
A (finite) \emph{trace} $\trace$ is a sequence of closed expressions
$\triple{e_0}{\vsu_0}{\tsu_0}, \ldots, \triple{e_n}{\vsu_n}{\tsu_n}$ such that
$\forall 0 \leq i < n$, we have
$\step{e_i}{\vsu_i}{\tsu_i}{e_{i+1}}{\vsu_{i+1}}{\tsu_{i+1}}$.
%
We write \steptr{\trace}{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'} if $\trace$ is
a trace of the form $\triple{e}{\vsu}{\tsu},\ldots,\pair{e'}{\vsu'}{\tsu'}$.
%
We write \steps{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'} if
\steptr{\trace}{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'} for some trace $\trace$.

@@

\paragraph{Witness Generality}
We now show that our evaluation relation instantiates holes in a
\emph{general} manner. That is, given a function $f$, if we have
$\steps{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,
then for every possible input type $t$, we can find a value
$\hastype{v}{t}$ such that $\eapp{f}{v}$ goes wrong.

@@

\paragraph{The Type of an Expression} The (dynamic) type of an
expression $e$ is defined as a function $\typeof{e}$ defined as:
  \[
  \begin{array}{lcll}
    \typeof{n}   & \defeq & \tint & \\
    \typeof{b}   & \defeq & \tbool & \\
    \typeof{\efun{x}{e}} & \defeq & \tfun \\
    \typeof{\vleaf{t}} & \defeq & \ttree{t} \\
    \typeof{\vnode{t}{v_1}{v_2}{v_3}} & \defeq & \ttree{t} \\
    \typeof{\vhole{t}} & \defeq & t \\
    \typeof{e} & \defeq & \thole, & \quad \text{\thole is fresh} \\
  \end{array}
  \]
%
Note that we approximate the type of expressions rather than
performing full type inference, for example
$\typeof{\efun{x}{\eplus{x}{1}}} = \tfun$,
instead of the usual $\tint \rightarrow \tint$.

@@

\paragraph{Type Compatibility}
Two types are compatible if one of them can be obtained by
suitably filling the holes of the other (\ie by suitably ``instantiating''
the type variables appearing in the other). Formally, a type $s$ is
\emph{compatible} with a type $t$, written \tcompat{s}{t}, if
$\exists \tsu.\ \subst{\tsu}{s} = \subst{\tsu}{t}$.

\paragraph{Preservation}
Given these two definitions, we show that each evaluation step
refines the partial type of $f$, \ie preserves type compatibility.
%
\begin{lem}
\label{lem:refine-partial}
If $\trace \defeq \triple{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu},\ldots$ and
$\trace' \defeq \trace, \triple{e'}{\vsu'}{\tsu'}$ (\ie $\trace'$ is a single-step
extension of $\trace$)
%
%The partial type of $f$ upto $\trace$ is compatible
%with the partial type upto $\trace'$, \ie\
%
then \tcompat{\ptype{\trace}{f}}{\ptype{\trace'}{f}}.
\end{lem}
\begin{proof}
  By induction on the length of $\trace$ and case analysis on the evaluation rules.
  %
  Note that all rules preserve partial types with the exception of when
  \forcesym\ is called on a hole, in which case we instantiate the hole with
  a concrete value.
  %
  But $\typeof{\ehole{}} = \thole{}$, which is compatible with any type.
\end{proof}

@@

\subsection{Search Algorithm}
\label{sec:search-algorithm}
%
So far, we have seen how a trace leading to a stuck configuration yields
a general witness demonstrating that the program is ill-typed (\ie\ goes
wrong for all inputs.)
In particular, we have shown how to non-deterministically find a witnesses
for a function of a \emph{single} argument.

>>>>

\paragraph{Holes}
\label{sec:holes}
%
Recall that a key challenge in our setting is to find witnesses
that are meaningful and do not arise from choosing values from
irrelevant types.
%
We solve this problem be equipping our term language with a notion
of a \emph{hole}, written \ehole, which represents \emph{unconstrained}
values that may be replaced with values from \emph{any} type.
%
Dually, holes may also appear in types, written \thole, where they may
be thought of as type variables that we will not generalize over.
%

\paragraph{Substitutions}
%
Our semantics ensure the generality of witnesses by incrementally
\emph{refining} holes, filling in just as much information as is
needed locally to make progress (inspired by the manner in
which SmallCheck uses lazy evaluation~\cite{runciman_smallcheck_2008}).
%
We track how the holes are incrementally filled in, by using
(value) \emph{substitutions} that map holes % (denoted by their index)
to values which may, in turn, contain holes.
%
The substitutions let us ensure that we instantiate each hole
with the same (partially defined) value, regardless of the multiple
contexts in which the hole appears, and hence, ensures we can
report a concrete (and general) witness for any (dynamically)
discovered type errors.

@@

\paragraph{Narrowing Types} The procedure % $\force{v}{t}{\vsu}{\tsu}$,
$$
\forcesym :: (v, t, \vsu, \tsu) \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu}
$$
defined in Figure~\ref{fig:narrow}, takes as input a value $v$, a type
$t$, and the current value and type substitutions, and refines $v$ to
have type $t$ by yielding a triple of either the same value and
substitutions, or yields the stuck state if no such refinement is
possible. In the case where $v$ is a hole, it first checks in the given
$\vsu$ to see if the hole has already been instantiated and, if so,
returns the existing instantiation.
\ES{explain the subtlety of narrowing a value hole to an uninstantiated type}
%
\begin{figure*}[ht]
$$
\begin{array}{lcl}
%% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
\forcesym                  & ::     & (v, t, \vsu, \tsu) \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu} \\
\force{v}{\thole}{\vsu}{\tsu}  & \defeq & \hspace{-1ex}
\begin{cases}
  \force{v}{\subst{\tsu}{\thole}}{\vsu}{\tsu} & \mbox{if}\ \thole \in dom(\tsu) \\
  \triple{v}{\vsu}{\extendsu{\tsu}{\thole}{\typeof{v}}} & \mbox{otherwise} \\
\end{cases} \\
\force{\vhole{t_1}}{t_2}{\vsu}{\tsu} & \defeq & \hspace{-1ex}
\begin{cases}
  \triple{v}{\vsu}{\tsu''}    & \mbox{if } v = \lookupsu{\vsu}{\ehole},
                                         \tsu' = \unify{t_1}{t_2}{\tsu},
                                         \tsu'' = \unify{\subst{\tsu'}{t_1}}
                                                        {\typeof{v}}{\tsu'} \\
  \triple{\stuck}{\vsu}{\tsu} & \mbox{if } v = \lookupsu{\vsu}{\ehole} \\
  \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu'} & \mbox{if}\ \tsu' = \unify{t_1}{t_2}{\tsu}, v = \gen{\subst{\tsu'}{t_1}}\\
\end{cases} \\
% \begin{cases}
%   \triple{\lookupsu{\vsu}{\ehole}}{\vsu}{\tsu} & \mbox{if}\ \ehole \in dom(\vsu), \hastype{\lookupsu{\vsu}{\ehole}}{t} \\
%   \triple{\stuck}{\vsu}{\tsu}                  & \mbox{if}\ \ehole \in dom(\vsu) \\
%   \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu}   & v = \gen{t}\\
% \end{cases} \\
 % \begin{array}{l}
    % \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \triple{\vsu(i)}{\vsu}\ \mathtt{else} \\
         % \elet{v}{\gen{t}}{\triple{v}{\ehole{i} \mapsto v}}
  % \end{array} \\
\force{n}{\tint}{\vsu}{\tsu}     & \defeq & \triple{n}{\vsu}{\tsu} \\
\force{b}{\tbool}{\vsu}{\tsu}    & \defeq & \triple{b}{\vsu}{\tsu} \\
\force{\efun{x}{e}}{\tfun}{\vsu}{\tsu} & \defeq & \triple{\efun{x}{e}}{\vsu}{\tsu} \\
\force{\vleaf{t_1}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vleaf{t_1}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{t_1}{t_2}{\tsu} \\
\force{\vnode{t_1}{v_1}{v_2}{v_3}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vnode{t_1}{v_1}{v_2}{v_3}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{t_1}{t_2}{\tsu} \\
\force{v}{t}{\vsu}{\tsu} & \defeq & \triple{\stuck}{\vsu}{\tsu}
\end{array}
$$
\caption{Narrowing values}
\label{fig:narrow}
\end{figure*}
%
While a hole may map to a value that \emph{contains} another hole, \eg a
lambda or a tree, it may not map \emph{directly} to another hole, so in
the second case of \forcesym\ we do not need to \forcesym\ the result of the
substitution, the sub-hole will be narrowed when the context demands it.

\paragraph{Generating Values} The (non-deterministic) procedure $\gen{t}$
takes as input a type $t$ and returns a value of that type. For base
types, the procedure returns an arbitrary value. For functions, it returns
a lambda with a \emph{new} hole denoting the return value, and unconstrained
types (denoted by $\thole$) yield fresh, unconstrained values
(denoted by $\ehole$).

@@

\paragraph{Steps and Traces}
\input{operational.tex}
%
% WRW notes that Figure 4 does not seem to handle recursion (it's not clear
% how the let rule would work for something "let rec"-y, and there's not
% function call rule). I only mention this because I can imagine a reviewer
% wondering about your ability to generate good witnesses for function
% types. This could likely be addressed in text, by a forward reference to
% Section 3.4 where higher-order functions are handled, without changing
% any of the formalisms at the last minute.
Figure~\ref{fig:operational} describes the small-step contextual
reduction semantics for \lang.
%
A \emph{configuration} is a triple $\triple{e \cup \stuck}{\vsu}{\tsu}$
of an expression $e$ or the stuck term $\stuck$, a value substitution
$\vsu$, and a type substitution $\tsu$.
%
We write $\step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$ if the configuration
$\triple{e}{\vsu}{\tsu}$ transitions in a \emph{small step} to
$\triple{e'}{\vsu'}{\tsu'}$.
%
A (finite) \emph{trace} $\trace$ is a sequence of configurations
$\triple{e_0}{\vsu_0}{\tsu_0}, \ldots, \triple{e_n}{\vsu_n}{\tsu_n}$ such that
$\forall 0 \leq i < n$, we have
$\step{e_i}{\vsu_i}{\tsu_i}{e_{i+1}}{\vsu_{i+1}}{\tsu_{i+1}}$.
%
We write \steptr{\trace}{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'} if $\trace$ is
a trace of the form $\triple{e}{\vsu}{\tsu},\ldots,\triple{e'}{\vsu'}{\tsu'}$.
%
We write \steps{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'} if
\steptr{\trace}{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'} for some trace $\trace$.

@@

\paragraph{Witness Generality}
We now show that our evaluation relation instantiates holes in a
\emph{general} manner. That is, given a function $f$, if we have
$\steps{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,
then for every possible input type $t$, we can find a value
$v$ of type $t$ such that $\eapp{f}{v}$ goes wrong.

@@

\paragraph{The Type of an Expression} The (dynamic) type of an
expression $e$ is defined as a function $\typeof{e}$ defined as:
  \[
  \begin{array}{lcll}
    \typeof{n}   & \defeq & \tint & \\
    \typeof{b}   & \defeq & \tbool & \\
    \typeof{\efun{x}{e}} & \defeq & \tfun \\
    \typeof{\vleaf{t}} & \defeq & \ttree{t} \\
    \typeof{\vnode{t}{v_1}{v_2}{v_3}} & \defeq & \ttree{t} \\
    \typeof{\vhole{t}} & \defeq & t \\
    \typeof{\eleaf} & \defeq & \ttree{\thole}, & \quad \text{\thole is fresh} \\
    \typeof{\enode{v_1}{v_2}{v_3}} & \defeq & \ttree{\thole}, & \quad \text{\thole is fresh} \\
    \typeof{e} & \defeq & \thole, & \quad \text{\thole is fresh} \\
  \end{array}
  \]
%
Note that we approximate the type of expressions rather than
performing full type inference, for example
$\typeof{\efun{x}{\eplus{x}{1}}} = \tfun$,
instead of the usual $\tint \rightarrow \tint$.

@@

\paragraph{Type Compatibility}
Two types are compatible if there exists a type subsitution that maps both
types to the same type.
% one of them can be obtained by
% suitably filling the holes of the other (\ie by suitably ``instantiating''
% the type variables appearing in the other).
%
Formally, a type $s$ is
\emph{compatible} with a type $t$, written \tcompat{s}{t}, if
$\exists \tsu.\ \subst{\tsu}{s} = \subst{\tsu}{t}$.

\paragraph{Preservation}
Given these two definitions, we show that each successful evaluation step
refines the partial type of $f$, \ie preserves type compatibility.
%
\begin{lem}
\label{lem:refine-partial}
If $\trace \defeq \triple{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu},\ldots$ and
$\trace' \defeq \trace, \triple{e'}{\vsu'}{\tsu'}$ (\ie $\trace'$ is a single-step
extension of $\trace$)
%
%The partial type of $f$ upto $\trace$ is compatible
%with the partial type upto $\trace'$, \ie\
%
then \tcompat{\ptype{\trace}{f}}{\ptype{\trace'}{f}}.
\end{lem}
\begin{proof}
  By induction on the length of $\trace$ and case analysis on the evaluation rules.
  %
  Note that all rules preserve partial types with the exception of when
  \forcesym\ is called on a hole, in which case we instantiate the hole with
  a concrete value.
  %
  But $\typeof{\ehole{}} = \thole{}$, which is compatible with any type.
\end{proof}

@@

\subsection{Search Algorithm}
\label{sec:search-algorithm}
%
So far, we have seen how a trace leading to a stuck configuration yields
a general witness demonstrating that the program is ill-typed (\ie\ goes
wrong for at least one input of every type.)
In particular, we have shown how to non-deterministically find a witnesses
for a function of a \emph{single} argument.
